Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-02-22 paperDistance/paper.tex
intro tweaks; notes to selves

With projects like \project{The Cannon} (DWH CITE) and \project{Avast}
(DWH CITE), we are exploring the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being answered.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ use case is ideal for thinking about purely statistical
models of stars:
The nearby stars are measured at very high signal-to-noise and can be
used to anchor empirical models that can be applied to the more
distant stars.

Here we deliver a demonstration-of-concept project that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise distances, without any input of the physics of stellar
evolution, interiors, or photospheres.
Most stellar models used to generate photometric distances or photometric
parallaxes have been physical models.
Because there are small issues with physical models of stars, leading to small
color and luminosity differences with the data, these models produce distance
estimates that are biased.
Besides, they build a long list of physical assumptions---assumptions about
how stars work and how to model them numerically---into the distance estimates.
That is, it is impossible (when using physical models)
to deliver distance estimates that involve minimal in assumptions.

However, the use of physical models for distance estimation is not necessary.
Because there are stars at all luminosities with good parallax information,
it is possible to build a stellar photometric model from the data themselves.
Different stars are observed at different levels of parallax precision,
so it requires relatively high technology to build this data-driven model
using all the data available, fairly and responsibly.

@@

>>>>

With projects like \project{The Cannon} (DWH CITE) and \project{Avast}
(DWH CITE), we are exploring the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ use case is ideal for thinking about purely statistical
models of stars:
The distance to a star is a simple latent variable, and the nearby
stars constitute a high signal-to-noise ``training set''.
These nearby stars can be used to anchor empirical models that can
then be applied to the more distant stars.

Here we deliver a demonstration-of-concept project that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise distances, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
Most stellar models used to generate photometric distances or photometric
parallaxes have been physical models, based on gravity, fluid mechanics,
radiative transfer, nuclear reactions, and atomic and molecular transitions.
Because there are small issues with each of these components, there are small
color and luminosity differences with the data---the models are wrong in detail---and
the physical models therefore produce distance estimates that are biased.
Besides, they build a long list of physical assumptions (about
nuclear reactions, convection, and thermal timescales, for example)
into the distance estimates.
That is, it is impossible, when using physical models,
to deliver photometric distance estimates that involve minimal assumptions.

However, the use of physical models for distance estimation is not necessary.
Because there are stars at all luminosities with good parallax information,
it is possible to build a stellar photometric model from the data themselves.
Different stars are observed at different levels of parallax precision,
so it requires relatively high technology to build this data-driven model
using all the data available, fairly and responsibly.
We have this technology!

@@

LA: Insert itemized or enumerated list of assumptions here. DWH will
typeset it to look okay! Be exhaustive and then we will edit down.
